---
title: "__PXD001283__: Mouse urinary bladder"
author: "Kylie Ariel Bemis"
date: "Revised: October 29, 2019"
output:
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{PXD001283: Mouse urinary bladder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

# Introduction

This vignette investigates the mouse urinary bladder dataset PXD001283 from PRIDE.

We reproduce some figures from the associate publication:

Römpp A, Guenther S, Schober Y, Schulz O, Takats Z, Kummer W, Spengler B; Histology by mass spectrometry: label-free tissue characterization obtained from high-accuracy bioanalytical imaging., Angew Chem Int Ed Engl, 2010 May 17, 49, 22, 3834-8, PMID: [20397170](http://europepmc.org/abstract/MED/20397170)

# Setup

First we set up our working environment.

We load the *Cardinal* package, register a serial backend (no parallelization), and specify a (parallelization-safe) random number generator.


```{r setup-1, message=FALSE}
library(Cardinal)
register(SerialParam())
RNGkind("L'Ecuyer-CMRG")
```

In addition, we turn off progress messages so they do not clutter the vignette.

```{r setup-2}
options(Cardinal.verbose=FALSE)
options(Cardinal.progress=FALSE)
```

# Files

Now, let's see what imzML files are available and create a scratch directory to save data if we need.

```{r files}
db <- "~/Documents/Datasets/PRIDE"

id <- "PXD001283"

datadir <- file.path(db, id)

tmpdir <- file.path(db, id, "tmp")

if ( !dir.exists(tmpdir) )
	dir.create(tmpdir)

filenames <- grep(".imzML", list.files(datadir), value=TRUE)
files <- file.path(datadir, filenames)
files
```

We have one imzML file.

# Import

We import the data using a 0.01 mz binwidth, as specified in the associated publication.

```{r read}
mse <- readMSIData(files, mass.range=c(400,1000),
					resolution=0.01, units="mz")
```

# Visualization (Fig. 1)

Below, we reproduce Fig. 1A from the publication.

```{r fig-1a}
image(mse, mz=c(741.5307, 798.5410, 743.5482),
		col=c("blue", "green", "red"),
		contrast.enhance="histogram",
		smooth.image="adaptive",
		normalize.image="linear",
		superpose=TRUE)
```

Our visualization looks very similar to the published figure.

# Mean spectrum

Below, we calcualate and plot the mean spectrum.

```{r mean}
mse_mean <- summarize(mse, .stat="mean")

plot(mse_mean)
```

Let's zoom in on two peaks shown in Fig. 2C which are only 0.05 *mz* apart.

```{r mean-zoom}
mse_mean %>%
	filter(mz >= 770.4, mz <= 770.7) %>%
	plot()
```

We see both peaks are visible in the mean spectrum.

# Pre-processing

We perform peak picking on the mean spectrum.

```{r peaks-1}
ref <- mse_mean %>%
	peakPick(method="mad") %>%
	peakAlign("mean",
			tolerance=0.01,
			units="mz") %>%
	peakFilter() %>%
	process()
```

Below, we see both peaks from Fig. 2C were detected.

```{r fig-2c}
mse_mean %>%
	filter(mz >= 770.4, mz <= 770.7) %>%
	plot()

abline(v=mz(ref), col="red", lty=2)
```

Now we bin the peaks in the full dataset to the reference peaks we detected in the mean spectrum.

```{r peaks-2}
mse_peaks <- mse %>%
	normalize(method="tic") %>%
	peakBin(ref=mz(ref),
			tolerance=0.01,
			units="mz") %>%
	process()
```

# Visualization (Fig. 2)

Below, we plot images for the two peaks at *mz* 770.51 and 770.56.

```{r fig-2b}
image(mse_peaks, mz=c(770.51, 770.56),
	contrast.enhance="histogram",
	smooth.image="adaptive",
	normalize.image="linear",
	layout=c(2,1))
```

We notice that these images are the reverse of what is shown in the paper (Fig. 2B), suggesting that these two peaks were mislabeled in the publication.

The color-coding in Fig. S5 from the Supporting Information suggests that this may be the case.


# Statistical analysis

Now we perform some statistical analysis, which was not done in the original paper.

We fit spatial shrunken centroids to segment the data.


```{r ssc}
set.seed(1)
ssc <- spatialShrunkenCentroids(mse_peaks,
			r=2, s=c(0,5,10,15,20), k=10,
			method="gaussian")

summary(ssc)

image(ssc, model=list(s=c(5,10,15,20)), key=FALSE)
```

We can see that the background takes up several segments, showing a left-to-right gradient across the slide.

We can see this same gradient in the clusters that make up the muscle layer, which is split into two segments.

When `s = 20`, the muscle segment becomes a single segment:

```{r}
image(ssc, model=list(s=20))
```

Below show the muscle layer (red, 2), the adventitial layer (green, 4), and the urothelium (blue, 5), which were detected.

```{r}
image(ssc, model=list(s=20), column=c(2,4,5))
```

Below, we plot the t-statistics that show the over/under-abundance of analytes in each of these segments relative to the mean spectrum.

```{r}
plot(ssc, values="statistic", model=list(s=20), column=c(2,4,5), lwd=2)
```

